Finding a closest point in a lattice of Voronoi's first kind

### 2014-01-13 paper.tex
added minor.  Needs a read through.

If $s$ is a short vector in a lattice $\Lambda$ then 
\[
\rho = \frac{\|s\|}{2} = \frac{1}{2} \min_{x \in \Lambda / \{\zerobf\} } \|x\|
\]
is called the \emph{packing radius} (or \emph{inradius}) of $\Lambda$~\cite{SPLAG}.  The packing radius is the minimum distance between the boundary of the Voronoi cell and the origin.  It is also the radius of the largest sphere that can be placed at every lattice point such that no two spheres intersect (see Figure~\ref{lattices:fig:vorregion}).  The following well known result will be useful.

@@

\begin{proposition}\label{eq:latticepointsinsphere}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice with packing radius $\rho$.  Let $S$ be an $m$-dimensional hypersphere of radius $r$ centered at $t \in \reals^{m}$.  The number of lattice points from $\Lambda$ in the sphere $S$ is less than $\ceil{r/\rho}^{n}$ where $\ceil{\cdot}$ denotes the smallest integer strictly larger than its argument. 
 \end{proposition}
 \begin{proof}
The packing radius $\rho$ is the Euclidean length of a point on the boundary of the Voronoi cell $\vor(\Lambda)$ that is closest to the origin. Therefore, the sphere $S$ is a subset of $\vor(\Lambda)$ scaled by $r/\rho$ and translated by $t$.  That is, $S \subset r/\rho \vor(\Lambda) + t$.  The proof follows because the number of lattice points in $r/\rho \vor(\Lambda) + t$ is less than $\ceil{r/\rho}^n$ (this follows from, for example, \cite[Theorem~2]{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}, see also~\cite[Lemma~3.7]{Micciancio09adeterministic}).
\end{proof}

@@

Two factors contribute to the computational complexity of this iterative approach to compute a closest lattice point.  The first factor is computing the minimum over the set $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer}.  In general a lattice can have as many as $2^{n+1}-2$ relevant vectors so computing a minimiser directly can require a number of operations that grows exponentially with $n$.  To add to this it is typically the case that the set of relevant vectors $\relevant(\Lambda)$ must be stored in memory so the algorithm can require an amount of memory that grows exponentially with $n$.  We will show that for a lattice of Voronoi's first kind the set of relevant vectors has a compact representation in terms of what is called its \emph{obtuse superbasis}.  To store the obtuse superbasis requires an amount of memory of order $O(n^2)$ in the worst case.  We also show that for a lattice of Voronoi's first kind the minimisation over $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer} can be solved efficiently by computing a minimum cut in an undirected flow network.  Using known algorithms a minimiser can be computed in $O(n^3)$ operations~\cite{Goldberg:1986:NAM:12130.12144,EdmondsKarp_max_flow,Cormen2001}.

@@

\begin{figure}[tp] 
	\centering      
		\includegraphics{figs/iterativeseriesexample-1.mps} 
		\caption{Example of the iterative procedure described in~\eqref{eq:relvectsminimslicer} to compute a closest lattice point to $y = (4,3.5)$ (marked with a cross) in the $2$-dimensional lattice generated by basis vectors $(2,0.4)$ and $(0.4,2)$.  The initial lattice point for the iteration is $x_0 = (-4.4,-2.8)$.  The shaded region is the Voronoi cell translated about the closest lattice point $x_6 = (4.4,2.8)$.}       
		\label{lattices:fig:iterativeexample} 
\end{figure}

@@

Classical examples of lattices of Voronoi's first kind are the $n$ dimensional root lattice $A_n$ and it's dual lattice $A_n^*$.  These lattices have convenient bases represented in $n+1$ dimensional space.  An obtuse superbasis for $A_n$ are $n+1$ vectors of length $n+1$ of the form
\begin{align*}
b_i &= e_i - e_{i+1}, \qquad i = 1, \dots, n \\
b_{n+1} &= e_{n+1} - e_{1},
\end{align*}
where $e_i$ is a vector of length $n+1$ with elements equal to zero except for the $i$th element that is equal to one.  The dual lattice $A_{n}^*$ is also of Voronoi's first kind.  The $i$th vector $b_i$ of the obtuse superbasis is of length $n+1$ with $j$th element equal to
\[
b_{ij} = \begin{cases}
\frac{n}{n+1} & j = i \\
-\frac{1}{n+1} & j \neq i.
\end{cases}
\]
For $A_n$ and $A_n^*$ there exists efficient algorithms that can compute a closest lattice point in $O(n)$ operations~\cite{McKilliam2009CoxeterLattices,Conway1982FastQuantDec}, and so, we do not recommend using the algorithm we describe here for these specific lattices.  %However, there exist a large number of lattice of first
%A lattice of Voronoi's first kind appears nontrivial.  One approach would be to construct the matrix $Q = B^\prime B$ containing the Selling parameters $q_{ij}$.  The matrix $Q$ requires to have nonpositive off diagonal entries and to be positive semidefinite with null space containing the vector $\onebf$ of all ones.  The explicit construction of such matrices is an interesting question for future research.  A related question is to determine, in some sense, how `large' the set of matrices satisfying these criteria is.  This would shed some light on how `large' the set of lattices of Voronoi's first kind is.

@@

The closest lattice point problem has a number of applications, for example, channel coding and data quantisation~\cite{Conway1983VoronoiCodes,Conway1982VoronoiRegions,Conway1982FastQuantDec,Erex2004_lattice_decoding,Erez2005}.  A significant hurdle in the practical application of lattices as codes or as quantisers is that computing a closest lattice point is computationally difficult in general~\cite{micciancio_hardness_2001}.  The best known general purpose algorithms require a number of operations of order $O(2^{2n})$~\cite{Micciancio09adeterministic}.

In this paper we have focused on the class of lattices of Voronoi's first kind.  We have shown that computing a closest point in a lattice of Voronoi's first kind can be achieved in a comparatively modest number of operations of order $O(n^4)$.  Besides being of theoretical interest, the algorithm has potential for practical application.  A question of immediate interest is: do there exist lattices of Voronoi's first kind that produce good codes or good quantisers?  Since lattices that produce good codes and quantisers often also describe dense~\emph{sphere packings}~\cite{SPLAG}, a related question is: do there exist lattices of Voronoi's first kind that produce dense sphere packings?  These question do not appear to have trivial answers.  The questions have heightened importance due to the algorithm described in this paper.

>>>>

If $s$ is a short vector in a lattice $\Lambda$ then 
\[
\rho = \frac{\|s\|}{2} = \frac{1}{2} \min_{x \in \Lambda / \{\zerobf\} } \|x\|
\]
is called the \emph{packing radius} (or \emph{inradius}) of $\Lambda$~\cite{SPLAG}.  The packing radius is the minimum distance between the boundary of the Voronoi cell and the origin.  It is also the radius of the largest sphere that can be placed at every lattice point such that no two spheres intersect (see Figure~\ref{lattices:fig:vorregion}).  The following well known results will be useful.

@@

\begin{proposition}\label{eq:latticepointsinvorcvell}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice.  For $r\in \reals$ let $\ceil{r}$ denote the smallest integer strictly larger than $r$.  Let $t \in \reals^m$.  The number of lattice points inside the scaled and translated Voronoi cell $r \vor(\Lambda) + t$ is atmost $\ceil{r}^n$.
\end{proposition}
\begin{proof}
It is convenient to modify the boundary of the Voronoi cell so that it tesselates $\reals^m$ under translations by $\Lambda$.  With this aim we let $V \subset \vor(\Lambda)$ contain all those points from the interior of $\vor(\Lambda)$, but have opposing faces open and closed.  That is, if $x \in V$ is on the boundary of $V$ then $-x \notin V$.  With this definition it can be asserted that $V$ tesslates $\reals^m$ under translations by $\Lambda$.  That is, $\reals^{m} = \cup_{x \in \Lambda}(V + x)$ and the intersection $(V + x)\cap(V+y)$ is empty for distinct lattice points $x$ and $y$~\cite[Sec.~2.2]{McKilliam2010thesis}.  Now, for positive integer $k$, the scaled and translated cell $kV + t$ contain precisely one of each of the coset representatives of the group quotient $\Lambda/k\Lambda$~\cite[Sec.~2.4]{McKilliam2010thesis}.  There are $k^n$ coset representatives.  Thus, the number of lattice point inside $r \vor(\Lambda) + t \subset \ceil{r}V + t$ is at most $\ceil{r}^n$.
\end{proof}

\begin{proposition}\label{eq:latticepointsinsphere}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice with packing radius $\rho$.  Let $S$ be an $m$-dimensional hypersphere of radius $r$ centered at $t \in \reals^{m}$.  The number of lattice points from $\Lambda$ in the sphere $S$ is less than $\ceil{r/\rho}^{n}$ where $\ceil{\cdot}$ denotes the smallest integer strictly larger than its argument. 
 \end{proposition}
 \begin{proof}
The packing radius $\rho$ is the Euclidean length of a point on the boundary of the Voronoi cell $\vor(\Lambda)$ that is closest to the origin. Therefore, the sphere $S$ is a subset of $\vor(\Lambda)$ scaled by $r/\rho$ and translated by $t$.  That is, $S \subset r/\rho \vor(\Lambda) + t$.  The proof follows because the number of lattice points in $r/\rho \vor(\Lambda) + t$ is less than $\ceil{r/\rho}^n$ by Proposition~\ref{eq:latticepointsinvorcvell}.
\end{proof}

@@

Two factors contribute to the computational complexity of this iterative approach to compute a closest lattice point.  The first factor is computing the minimum over the set $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer}.  In general a lattice can have as many as $2^{n+1}-2$ relevant vectors so computing a minimiser directly can require a number of operations that grows exponentially with $n$.  To add to this it is often the case that the set of relevant vectors $\relevant(\Lambda)$ must be stored in memory so the algorithm can require an amount of memory that grows exponentially with $n$~\cite[Sec.~6]{Micciancio09adeterministic}\cite{Shalvi_iterativeslicer_2009}.  We will show that for a lattice of Voronoi's first kind the set of relevant vectors has a compact representation in terms of what is called its \emph{obtuse superbasis}.  To store the obtuse superbasis requires an amount of memory of order $O(n^2)$ in the worst case.  We also show that for a lattice of Voronoi's first kind the minimisation over $\relevant(\Lambda) \cup \{\zerobf\}$ in~\eqref{eq:relvectsminimslicer} can be solved efficiently by computing a minimum cut in an undirected flow network.  Using known algorithms a minimiser can be computed in $O(n^3)$ operations~\cite{Goldberg:1986:NAM:12130.12144,EdmondsKarp_max_flow,Cormen2001}.

@@

\begin{figure}[tp] 
	\centering      
		\includegraphics{figs/iterativeseriesexample-1.mps} 
		\caption{Example of the iterative procedure described in~\eqref{eq:relvectsminimslicer} to compute a closest lattice point to $y = (4,3.5)$ (marked with a cross) in the $2$-dimensional lattice generated by basis vectors $(2,0.4)$ and $(0.4,2)$.  The initial lattice point for the iteration is $x_0 = (-4.4,-2.8)$.  The shaded region is the Voronoi cell surrounding the closest lattice point $x_6 = (4.4,2.8)$.}       
		\label{lattices:fig:iterativeexample} 
\end{figure}

@@

Classical examples of lattices of Voronoi's first kind are the $n$ dimensional root lattice $A_n$ and its dual lattice $A_n^*$~\cite{SPLAG}.  These lattices have convenient bases represented in $n+1$ dimensional space.  An obtuse superbasis for $A_n$ is of the form
\begin{align*}
b_i &= e_i - e_{i+1}, \qquad i = 1, \dots, n \\
b_{n+1} &= e_{n+1} - e_{1},
\end{align*}
where $e_i$ is a vector of length $n+1$ with elements equal to zero except for the $i$th element that is equal to one.  The dual lattice $A_{n}^*$ is also of Voronoi's first kind.  The $i$th vector $b_i$ of the obtuse superbasis has $j$th element equal to
\[
b_{ij} = \begin{cases}
\frac{n}{n+1} & j = i \\
-\frac{1}{n+1} & j \neq i.
\end{cases}
\]
For $A_n$ and $A_n^*$ there exists efficient algorithms that can compute a closest lattice point in $O(n)$ operations~\cite{McKilliam2009CoxeterLattices,Conway1982FastQuantDec} so we do not recommend using the algorithm described in this paper for $A_n$ and $A_n^*$.  The fast algorithms for $A_n$ and $A_n^*$ exploit the special structure of these lattices and are not applicable to other lattices.  In contrast, the algorithm we describe here works for all lattices of Voronoi's first kind.

%However, there exist a large number of lattice of first
%A lattice of Voronoi's first kind appears nontrivial.  One approach would be to construct the matrix $Q = B^\prime B$ containing the Selling parameters $q_{ij}$.  The matrix $Q$ requires to have nonpositive off diagonal entries and to be positive semidefinite with null space containing the vector $\onebf$ of all ones.  The explicit construction of such matrices is an interesting question for future research.  A related question is to determine, in some sense, how `large' the set of matrices satisfying these criteria is.  This would shed some light on how `large' the set of lattices of Voronoi's first kind is.

@@

The closest lattice point problem has a number of applications, for example, channel coding and data quantisation~\cite{Conway1983VoronoiCodes,Conway1982VoronoiRegions,Conway1982FastQuantDec,Erex2004_lattice_decoding,Erez2005}.  A significant hurdle in the practical application of lattices as codes or as quantisers is that computing a closest lattice point is computationally difficult in general~\cite{micciancio_hardness_2001}.  The best known general purpose algorithms require a number of operations of order $O(2^{2n})$~\cite{Micciancio09adeterministic}.  In this paper we have focused on the class of lattices of Voronoi's first kind.  We have shown that computing a closest point in a lattice of Voronoi's first kind can be achieved in a comparatively modest number of operations of order $O(n^4)$.  Besides being of theoretical interest, the algorithm has potential for practical application.

 It is straightforward to construct an `arbitrary' lattice of Voronoi's first kind.  One approach is to construct the $n+1$ by $n+1$ symmetric matrix $Q = B^\prime B$ with elements $Q_{ij} = q_{ij} = \dotprod{b_{i}}{b_j}$ given by the Selling parameters.  Choose the off diagonal entries of $Q$ to be nonpositive with $q_{ij}=q_{ji}$ and set the diagonal elements $q_{ii} = -\sum_{j=1}^{n+1} q_{ij}$.  %We then have $Q\onebf = \zerobf$ and so $\onebf$ is in the null space of $Q$.  
The matrix $Q$ is diagonally dominant, that is, $\abs{q_{ii}} \geq \sum_{j=1}^{n+1} \abs{q_{ij}}$, and so $Q$ is positive semidefinite.  A rank deficient Cholesky decomposition~\cite{Higham90analysisof} can now be used to recover a matrix $B$ such that $B^\prime B  = Q$. The columns of $B$ are vectors of the obtuse superbasis.

A question of immediate interest is: do there exist lattices of Voronoi's first kind that produce good codes or good quantisers?  Since lattices that produce good codes and quantisers often also describe dense~\emph{sphere packings}~\cite{SPLAG}, a related question is: do there exist lattices of Voronoi's first kind that produce dense sphere packings?  These question do not appear to have trivial answers.  The questions have heightened importance due to the algorithm described in this paper.
